paper2_resistance_mechanisms_mixtures

### 2016-09-12 inst/documents/paper2_resistance_mechanisms_mixtures.Rmd
paper writing

(This file is automagically generated by R from paper2_resistance_mechanisms_mixtures.Rmd)
Target : Malaria journal

#### TO DO
1. decide whether to use selective advantage or rr restoration coefficient

@@

The simulation represents a population of randomly mixing individuals using standard population genetic approaches so that it does not need to follow every individual. One locus is represented for each insecticide. There are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. The fitness of an individual depends on it's alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide, the dominance of the resistance allele and the ability of resistance to restore fitness in the presence of the insecticide (see Table 1).

@@

So far this description has just considered a single insecticide and associated resistance allele (Fig 1A). In the simulation a secound insecticide and resistance allele are included and fitness is calculated by multiplying these with those for the first insecticide described above (Fig 1C).

In addition the model structure allows for 3 different levels of exposure to be set for each insecticide (Fig 2). So in the two insecticide case that gives the potential for nine (3*3) different exposure niches. In the runs shown here we have just used two levels of exposure for each insecticide (effectively no or yes with the value for yes set from the exposure input in Table 1). The model structure allows males and females to experience different exposures.

@@

For single insecticide use, faster resistance resulted from higher values of any of the inputs outlined in Table 1, namely effectiveness, exposure, dominance, the resistance restoration coefficient, and starting frequencies of resistance (Figs 3-5). In the plots each input was varied in isolation with the remaining set at a constant 0.5 (except for starting frequency of resistance which was set to 0.01).

When two insecticides are used the relative performance of a mixture strategy compared to sequential use is determined by the values of these inputs. To help us understand the mechanisms influencing this relative performance we started with a single base scenario and investigated changing inputs indivividually. For our base scenario we set the values for the first four inputs in Table 1 to 0.5 and the starting frequencies of resistance to 0.1. For this base scenario resistance arises slower for sequential than mixture (Fig 6A). This plot, and those that follow, compares mixture and sequence for a single input scenario (where the previous plots showed 3 input scenarios). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time. When used in sequence the curve (dashed) for each insecticide individually is steeper but, because they happen one after the other, it takes longer to reach the resistance threshold for both.

### Insecticide effectiveness
A more effective insecticide prompts increased time-to-resistance when used in a mixture. This contrasts with the earlier result showing that when used in isolation a single insecticide prompts a shorter time-to-resistance.

For example taking the base scenario and increasing the effectiveness of one insecticide from 0.5 to 0.8 increases the time-to-resistance for the mixture and decreases that for the sequence (Fig 6B). In the mixture resistance to insecticide1 (with the increased effectiveness) rises at a similar speed to the base scenario (Fig 6A).

When the effectiveness of both insecticides are increased to 0.8 the time-to-resistance is increased still further to around 125 generations (Fig 7). In this case the curves for the rise in resistance to each insecticide are identical. The curves are shallower than previous plots because individuals that are resistant to each insecticide are being killed by the other thus reducing the increase in resistance alleles. As the resistance frequencies go above 10% around generation 100 both insecticides provide less protection to the other and the slope of the resistance curve increases.

### Exposure
Increasing exposure to 0.8 reduces time-to-resistance for both mixture and sequential scenarios (Fig 8A relative to 4A). The change in exposure has a greater effect on the mixture relative to the sequential strategy and this results in a greater relative benefit for sequence versus mixture. Remembering that earlier increasing one effectiveness to 0.8 resulted in mixtures being favoured over sequence (Fig 6B). Increasing exposure from 0.5 to 0.8 switches back to sequential being the favoured strategy (Fig 8B). This demonstrates how higher exposures can favour sequential use.

### Dominance of resistance alleles
Increasing the dominance of the alleles within a mixture has the same direction of effect as it does for sole use. Increasing the dominance of the allele corresponding to insecticide1 from 0.5 to 0.8 decreases the time-to-resistance to insecticide1 within the mixture but leaves the time-to-resistance for the second insecticide in the mixture the same (Fig 9A). Setting the dominance for both alleles to 0.8 reduces the time-to-resistance for both the mixture and the sequence by 20 generations from the base scenario (compare Fig 9B to 6A). In both of these cases sequential use continues to have a longer time-to-resistance under the higher dominance.

### Resistance restoration coefficient
Increasing the resistance restoration coefficient to insecticide1 from 0.5 to 0.8 decreases the time-to-resistance for insecticide1 in the mixture, and has a similar effect on it's use in the sequence (Fig 10A). However, it has little effect on resistance to insecticide2. Because of this sequential use continues to generate a longer time-to-resistance than the mixture. Increasing the resistance restoration coefficient of both alleles to 0.8 results in a shorter time-to-resistance  for the mixture (Fig 10B), however the relative shape of the resistance curves remains similar to the base scenario  (Fig 6A), with sequential use continuing to generate longer time-to-resistance.

### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 11A to Fig 6A). Similarly taking the base scenario with effectiveness of the first insecticide increased to 0.8 (Fig 6B) and decreasing starting frequency of that resistance allele did not result in the mixture being favoured (Fig 11B).

@@

When single insecticide use was represented (Figs 3-5) higher values of all of the inputs led to faster resistance. 
[write more about single insecticide responses here]

The response of resistance to changing exposure to, and effectiveness of, the insecticide was identical (compare Fig 3A to 3B). This makes sense given, for example, that exposing half of a vector population to an insecticide which kills a quarter of the exposed would be expected to kill the same number as exposing quarter of the population to an insecticide that kills a half of the exposed. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

These mechanisms explaining the response of resistance to single insecticide use are summarised in Table 2.

In comparison with the use of a single insecticide, using two insecticides either in sequence or in mixtures differs in some ways and where it differs that can again help us to understand the mechanisms.

[explain the effectivenes result, referring to Fig 4B versus 4A]
In contrast, resistance to insecticide2 increases more slowly initially. Once insecticide1 reaches it's resistance threshold, the curve for insecticide2 becomes steeper. This can be explained by the mechanism that once insecticide1 has become ineffective due to resistance it stops killing indivduals that are resistant to insecticide2 and thus stops 'protecting' it from it's own rise in resistance.

@@

(@) For a mixture of 2 insecticides with differing effectiveness. Resistance to the insecticide with a greater effectiveness increases faster. The less effective insecticide is 'protected' by the more effective. Resistance to the less effective insecticide increases slowly until resistance to the more effective insecticide reaches a high level at which point it starts to increase faster.

## References
1. GPIRM
1. Curtis 
1. Mani GS. Evolution of resistance in the presence of two insecticides. Genetics 1985;109(4):761–83.
1. Birget 2015
1. Curtis et al 1993, potential IRM strategies :
~ moderation : preserve susceptible genes by limiting selection pressure
~ saturation : high dose so that heterozygous resistants are killed
~ multiple attack : independently acting pressures neither of which is strong enough to lead to resistance

@@

1. work out what's going on with exposure ... why does increasing exposure have a greater effect on mixture than sequential ? Is it because the exposure to one insecticide also effects the development of resistance in the other ?
1. In discussion and maybe elewhere address potential concern that the relationships shown here may not hold for all of paramter space.
1. re-read GPIRM
1. can I develop explanatory diagrams to describe the input parameters ?
1. mention generation time of 1 month
1. mention other potential effects e.g. differential exposure of M&F, sex linkage, ...

@@

![Building up the model structure.](mod_graphic_a_to_c.png)

![Exposure to insecticides in the model.](mod_graphic_d.png)

@@

>>>>

Target : Malaria journal

THIS IS A CONFIDENTIAL COPY OF THE MS. Please do not circulate without the express permission of one of the authors. All comments and & suggestions welcome.

(This file is automagically generated by R from paper2_resistance_mechanisms_mixtures.Rmd)

## ANDY TO DO
1. work out what's going on with exposure ... why does increasing exposure have a greater effect on mixture than sequential ? Is it because the exposure to one insecticide also effects the development of resistance in the other ?
1. check & add to discussion text
1. add to intro about results in levick et al. 
1. explain resistance-restoration and its relation to selection coefficient

@@

The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One locus is represented for each insecticide. There are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. The fitness of an individual depends on it's alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide, the dominance of the resistance allele and the ability of resistance to restore fitness in the presence of the insecticide (see Table 1).

@@

So far this description has just considered a single insecticide and associated resistance allele (Fig 1A). In the simulation a secound insecticide and resistance allele are included and fitness is calculated by multiplying these with those for the first insecticide described above (Fig 1C). Thus the simulation can represent populations exposed to two insecticides together in a mixture.

In addition the model structure allows for 3 different levels of exposure to be set for each insecticide (Fig 2). So in the two insecticide case that gives the potential for nine (3*3) different exposure niches. In the runs shown here we have just used two levels of exposure for each insecticide (effectively no or yes with the value for yes set from the exposure input in Table 1). The model structure allows males and females to experience different exposures, we explore that in Levick et al. (in press) but the runs presented here use equal exposures.

Here we principally look at model outputs of the change in allele frequencies over time (generations). We investigate three insecticide use strategies.
1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the proportion of resistance alleles reaches 50%
1. two insecticides used in a mixture (concurrently)

@@

### Single insecticide
For single insecticide use, faster resistance resulted from higher values of any of the inputs outlined in Table 1, namely effectiveness, exposure, dominance, the resistance restoration coefficient, and starting frequencies of resistance (Figs 3-5). In the plots each input was varied in isolation with the remaining set at a constant 0.5 (except for starting frequency of resistance which was set to 0.01).

### Two insecticides
When two insecticides are used the relative performance of a mixture strategy compared to sequential use changes dependent on input values. To help us understand the mechanisms influencing this relative performance we started with a single base scenario and investigated changing inputs individually. For our base scenario we set the values for the first four inputs in Table 1 to 0.5 and the starting frequencies of resistance to 0.1. For this base scenario resistance arises slower for sequential than mixture (Fig 6A). This plot, and those that follow, compares mixture and sequence for a single input scenario (where the previous plots showed 3 input scenarios). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer to reach the resistance threshold for both.

#### results: Insecticide effectiveness
A more effective insecticide prompts increased time-to-resistance when used in a mixture. This is opposite to what happens when used in sequence and in isolation, both prompting a shorter times-to-resistance.

For example, taking the base scenario and increasing the effectiveness of one insecticide from 0.5 to 0.8 increases the time-to-resistance for the mixture (from 90 to 115 generations) and decreases that for the sequence (from 120 to 90 generations) (Fig 6). Thus at low effectiveness a sequential strategy is favoured while at higher effectiveness a mixture is favoured.

*? it would be nice to add grey lines to fig6 & others for the 'other' scenario to facilitate comparison, but I'd need to do some tricky rewriting of Beths code.*

When the effectiveness of both insecticides are increased to 0.8 the time-to-resistance is increased still further to around 125 generations (Fig 7).

#### results: Exposure
Increasing exposure to 0.8 reduces time-to-resistance for both mixture and sequential scenarios (Fig 8A relative to 4A). The change in exposure has a greater effect on the mixture relative to the sequential strategy and this results in a greater relative benefit for sequence versus mixture. Remembering that earlier increasing one effectiveness to 0.8 resulted in mixtures being favoured over sequence (Fig 6B). Increasing exposure from 0.5 to 0.8 switches back to sequential being the favoured strategy (Fig 8B). This demonstrates how higher exposures can favour sequential use.

#### results: Dominance of resistance alleles
Increasing the dominance of the alleles within a mixture has the same direction of effect as it does for sole use. Increasing the dominance of the allele corresponding to insecticide1 from 0.5 to 0.8 decreases the time-to-resistance to insecticide1 within the mixture but leaves the time-to-resistance for the second insecticide in the mixture the same (Fig 9A). Setting the dominance for both alleles to 0.8 reduces the time-to-resistance for both the mixture and the sequence by 20 generations from the base scenario (compare Fig 9B to 6A). In both of these cases sequential use continues to have a longer time-to-resistance under the higher dominance.

#### results: Resistance restoration coefficient
Increasing the resistance restoration coefficient to insecticide1 from 0.5 to 0.8 decreases the time-to-resistance for insecticide1 in the mixture, and has a similar effect on it's use in the sequence (Fig 10A). However, it has little effect on resistance to insecticide2. Because of this sequential use continues to generate a longer time-to-resistance than the mixture. Increasing the resistance restoration coefficient of both alleles to 0.8 results in a shorter time-to-resistance  for the mixture (Fig 10B), however the relative shape of the resistance curves remains similar to the base scenario  (Fig 6A), with sequential use continuing to generate longer time-to-resistance.

#### results: Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 11A to Fig 6A). Similarly taking the base scenario with effectiveness of the first insecticide increased to 0.8 (Fig 6B) and decreasing starting frequency of that resistance allele did not result in the mixture being favoured (Fig 11B).

@@

When single insecticide use was represented (Figs 3-5) in all cases the initial increase in resistance was most rapid with straight lines on the log plots up to around 50% followed by a slowing rate of increase up to 100%. Higher values of the five inputs invetsigated all led to more rapid increases in resistance.

The response of resistance to changing exposure and effectiveness was identical (compare Fig 3A to 3B). This makes sense as vector kill is effectively a product of exposure times effectiveness. For example exposing 50% of a  population to an insecticide which is 75% effective would be expecetd to have the same result as exposing 75% of the population to one which is 50% effective. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

*can we explain why at low dominance 0.2 the rate of increase in resistance is slower at the start ?*

The pattern of more rapid increase in resistance at higher dominance levels (Fig 4A) can be explained by increased survival of heterozygotes in the presence of the insecticide. Higher dominance of the resistant allele causes it to contribute more to the phenotype of the heterozygotes leading to higher survival. Thus selection pressure for the resistance allele will be increased because it confers more of an advantage when only present on one chromosome.

The faster development of resistance under higher resistance-restoration (Fig 4B) can be explained by it's effect on the survival of the resistant genotypes. Resistance-restoration restores the survival of resistant genotypes in the presence of the insecticide back towards what it would be in the absence of the insecticide, thus increasing the selection pressure of the resistance allele.

The effect of the starting frequency of resistance (Fig 5) is the most different from the other inputs and perhaps the easiest to explain. Simply when starting from a higher frequency of resistance there is a smaller change to make to reach the higher resistance thresholds.

These mechanisms explaining the response of resistance to single insecticide use under different input values are summarised in Table 2 and Fig 12. Exposure and effectiveness act by decreasing the survival of all genotypes, dominance acts by increasing survival of RS and resistance-restoration acts by increasing survival of RR & RS.

#### discussion: Two insecticides

In comparison with the use of a single insecticide, using two insecticides either in sequence or in mixtures differs in some ways and where it differs that can again help us to understand the mechanisms. Looking at our base scenario (Fig 6A)

#### discussion: Insecticide effectiveness

The main difference between mixtures and the other strategies is that in mixtures increasing the effectiveness of either insecticide slows the development of resistance to both (Fig 6B). In contrast in both the sequential (Fig 6B) and single insecticide (Fig 3A) experiments resistance was speeded up by increasing effectiveness. This can be explained by the effect that in a mixture increased effectiveness of one insecticide increases killing of vectors that are resistant to the other and thus decreases selection pressure.

This effect can be seen by comparing Fig 6B, where the effectiveness of insecticide 1 has been increased, to 6A. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. This can be explained by the mechanism that once insecticide 1 has become ineffective due to resistance it stops killing indivduals that are resistant to insecticide 2 and thus stops 'protecting' it from it's own rise in resistance. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 7). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 6A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 6B.

Fig 6B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal resistance will be expected to increase faster to the more effective insecticide. The more effective insecticide prompts both a) higher selection pressure for it's own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide.

#### discussion: Exposure

*may want to copy or move some text from summary points below into discussion here*

### Discussion summary points

(@) For a mixture of 2 insecticides with differing effectiveness. Resistance to the insecticide with a greater effectiveness increases faster. The less effective insecticide is 'protected' by the more effective. Resistance to the less effective insecticide increases slowly until resistance to the more effective insecticide reaches a high level.

@@

## References
1. GPIRM
1. Curtis 
1. Mani GS. Evolution of resistance in the presence of two insecticides. Genetics 1985;109(4):761–83.
1. Birget 2015
1. Curtis et al 1993, 
1. Georghiou 1983 Management of resistance in Arthropods.
potential IRM strategies :
~ moderation : preserve susceptible genes by limiting selection pressure
~ saturation : high dose so that heterozygous resistants are killed
~ multiple attack : independently acting pressures neither of which is strong enough to lead to resistance

@@

1. In discussion and maybe elewhere address potential concern that the relationships shown here may not hold for all of paramter space.
1. can I develop explanatory diagrams to describe the input parameters ?
1. mention generation time of 1 month
1. mention other potential effects e.g. differential exposure of M&F, sex linkage, ...
1. add to discussion about relevance of these mechanisms in the field
1. e.g. exposure can be reduced if either insecticide has a repellant effect. (Birget & Koella 2015 mention this not for resistance but for ITNs with repellancy requiring higher coverage for elimination)

@@

![Building up the model structure.](paper2_mod_graphic_a_to_c.png)

![Exposure to insecticides in the model.](paper2_mod_graphic_d.png)

@@

![Summarising mechanisms of single insecticide effects.](paper2_input_effects_single.png)
